---
title: "Differential expression analysis"
author: "Marco Tello"
date: "2024-02-14"
output: html_document
---

```{r setup, include=FALSE}
# Differential expression tools
library(DESeq2)
# Data format and ID conversion
library(data.table)
library(biomaRt)
library(stringr)
# Plot generation 
library(ggplot2)
library(factoextra)
library(pheatmap)

theme_set(theme_bw())
```


```{r custom functions, include=FALSE}
# Custom heatmap function
custom_volcano <- function(DEG_results, alpha, log2FC){
  temp <- data.table(na.omit(DEG_results))
  temp[, alpha := ifelse(padj <= alpha, TRUE, FALSE)]
  temp[, log2FC := ifelse(abs(log2FoldChange) >= log2FC, TRUE, FALSE)]
  temp[, DEG := "No"]
  temp[, DEG := ifelse(alpha & !log2FC, "FDR", DEG)]
  temp[, DEG := ifelse(alpha & log2FC, "FDR and FC", DEG)]
  ggplot(temp, aes(x = log2FoldChange, y = -log10(padj), color = DEG)) +
    geom_hline(yintercept = -log10(alpha), linetype = "dashed", color = "gray") +
    geom_vline(xintercept = log2FC, linetype = "dashed", color = "gray") +
    geom_vline(xintercept = -log2FC, linetype = "dashed", color = "gray") +
    geom_point(alpha = 0.4, size = 3) +
    # xlim(-6, 3)+
    scale_color_manual(values = c("No" = "darkgray", "FDR" = "blue", "FDR and FC" = "red")) +
    labs(x = "log2( fold change )", y = "-log10( adjusted p-value)", color = "DEG status")
}

```


```{r gene conversion table}
conversion_table <- fread("~/GitHub/CellDecon/input/rnaseq/conversion_table.tsv")

```



# Standard DEA

First we perform differential expression analysis utilizing only the differences between diets to identify DEGs. 

```{r read data}

# Read expression data
exp_df <- fread(file = "~/GitHub/CellDecon/input/rnaseq/filtered_counts.tsv")
rna_mat <- as.matrix(exp_df[,3:ncol(exp_df)])
rownames(rna_mat) <- exp_df$ENSEMBLID

# Build metadata
condition <- str_replace(colnames(rna_mat), pattern = ".\\d+", replacement = "") 
condition <- str_replace(condition, pattern = "_\\d{6}.\\d+", replacement = "") 
condition <- str_replace(condition, pattern = "CSAAPTS", replacement = "PTS") 

coldat <- data.frame(Diet = as.factor(condition),
                     Sample = colnames(rna_mat), 
                     stringsAsFactors = TRUE)
coldat$Diet <- relevel(coldat$Diet, ref = "CSAA")

```


Use DESeq2 to identify differences between diets 

```{r}
deseq_rna_base <- DESeqDataSetFromMatrix(countData = rna_mat,
                                         colData = coldat,
                                         design = ~ Diet)

deseq_rna_base <- DESeq(deseq_rna_base)

RNA_DE_PTS_vs_CSAA <- results(deseq_rna_base, 
                              name="Diet_PTS_vs_CSAA",
                              pAdjustMethod = "BH", 
                              alpha = 0.05)
RNA_DE_PTS_vs_CSAA <- cbind(rownames(RNA_DE_PTS_vs_CSAA),
                            RNA_DE_PTS_vs_CSAA)
names(RNA_DE_PTS_vs_CSAA)[1] <- "Gene"
RNA_DE_PTS_vs_CSAA <- as.data.table(na.omit(RNA_DE_PTS_vs_CSAA))
RNA_DE_PTS_vs_CSAA[, padj := round(padj, 2)]

out_file <- "~/GitHub/CellDecon/output/DESeq2/base_DESeq_full.tsv"
if(!file.exists(out_file)){
  fwrite(RNA_DE_PTS_vs_CSAA, file = out_file, 
         append = FALSE, quote = FALSE, 
         sep = '\t', na = "",
         row.names = FALSE, col.names = TRUE)
}
```

## Visualize results

```{r}
hist(RNA_DE_PTS_vs_CSAA$pvalue, main = "CSAA vs PTS", xlab = "p-value")

alpha = 0.05
log2FC = 1
custom_volcano(RNA_DE_PTS_vs_CSAA, alpha, log2FC)

```

```{r}

base_DEGs <- merge.data.table(x = conversion_table,
                              y =  RNA_DE_PTS_vs_CSAA, 
                              by.x = "ENSEMBL", 
                              by.y = "Gene")

base_DEGs[, padj := round(padj, 2)]

out_file <- "~/GitHub/CellDecon/output/DESeq2/standard_DEG.tsv"
if(!file.exists(out_file)){
  fwrite(base_DEGs[padj <= 0.05], file = out_file, 
         append = FALSE, quote = FALSE, 
         sep = '\t', na = "",
         row.names = FALSE, col.names = TRUE)
}

```



# Cell deconvolution results


First, we read the calculated cell fractions. 

```{r}

fractions <- fread("~/GitHub/CellDecon/output/CIBERSORTx/fractions_Sone2one_Mone2one_newExp.txt")
names(fractions) <- str_replace(str_replace(names(fractions), pattern = " Lineage", replacement = ""),pattern = " ", replacement = "")
fractions <- fractions[, .SD, .SDcols = c("Mixture","BCell","TCell","Macrophage","Neutrophil")]
fractions[, Mixture := gsub(pattern = "-", replacement = ".", x = Mixture)]
fractions <- data.table(fractions, key = "Mixture")

# Verify that samples are in the same order as the input rna_mat
fractions_sub <- fractions[colnames(rna_mat)]
fractions_norm <- fractions_sub[, .SD/sum(.SD), .SDcols = !"Mixture", by = Mixture]

# Format sample metadata 
coldat_decon <- merge.data.table(x = coldat, y = fractions_sub, 
                                 by.x = "Sample", by.y = "Mixture")
coldat_decon$Diet <- relevel(coldat_decon$Diet, ref = "CSAA")
```

Then we identify the differentially expressed genes incorporating the cell fractions. 

```{r}
# Build DESeq object to perform the analysis
deseq_rna_decon <- DESeqDataSetFromMatrix(countData = rna_mat,
                                          colData = coldat_decon,
                                          design = ~ Diet + TCell + Macrophage + BCell )
deseq_rna_decon <- DESeq(deseq_rna_decon)

decon_RNA_DE_PTS_vs_CSAA <- results(deseq_rna_decon, 
                                    name="Diet_PTS_vs_CSAA",
                                    pAdjustMethod = "BH", 
                                    alpha = 0.05)
decon_RNA_DE_PTS_vs_CSAA <- cbind(rownames(decon_RNA_DE_PTS_vs_CSAA),
                                  decon_RNA_DE_PTS_vs_CSAA)
names(decon_RNA_DE_PTS_vs_CSAA)[1] <- "Gene"
decon_RNA_DE_PTS_vs_CSAA <- as.data.table(na.omit(decon_RNA_DE_PTS_vs_CSAA))

out_file <- "~/GitHub/CellDecon/output/DESeq2/decon_DESeq_full.tsv"
if(!file.exists(out_file)){
  fwrite(decon_RNA_DE_PTS_vs_CSAA, file = out_file, 
         append = FALSE, quote = FALSE, 
         sep = '\t', na = "",
         row.names = FALSE, col.names = TRUE)
}

```


## Visualize deconvolution results 

```{r}
hist(decon_RNA_DE_PTS_vs_CSAA$pvalue, main = "CSAA vs PTS", xlab = "p-value")

alpha = 0.05
log2FC = 1
custom_volcano(decon_RNA_DE_PTS_vs_CSAA, alpha, log2FC)
```

```{r}
decon_DEGs <- merge.data.table(conversion_table, 
                               decon_RNA_DE_PTS_vs_CSAA, 
                               by.x = "ENSEMBL", 
                               by.y = "Gene")

decon_DEGs[, padj := round(padj, 2)]

out_file <- "~/GitHub/CellDecon/output/DESeq2/decon_DEG.tsv"
if(!file.exists(out_file)){
  fwrite(decon_DEGs[padj < 0.05], file = out_file, 
         append = FALSE, quote = FALSE, 
         sep = '\t', na = "",
         row.names = FALSE, col.names = TRUE)
}

```








